---
title: "*Cepaea nemoralis* snails are less likely to trap nematodes within their shells in urban areas - main analysis"
author: "Maxime Dahirel"
format: html
editor_options: 
  chunk_output_type: console
---

```{r packages}
library(car)       # CRAN v3.1-2
library(DHARMa)    # CRAN v0.4.6
library(emmeans)   # CRAN v1.9.0
library(glmmTMB)   # CRAN v1.1.8
library(MuMIn)     # CRAN v1.47.5
library(ncf)       # CRAN v1.3-2
library(sf)        # CRAN v1.0-15

library(tidyverse) # CRAN v2.0.0

library(here)      # CRAN v1.0.1
```

# Loading data

We first load our snail data and our environmental data:

```{r data}
nematodes_raw <- read_csv(here("data","HELICITY22_nematodes.csv"))
colours_raw <- read_csv(here("data","HELICITY22_shellcolours.csv"))

points_raw <- st_read(here("data","HELICITY22_sites_BELambert72_wGHSL.gpkg")) 
```

The `nematodes_raw` dataset is a simple data frame with three columns:

- `Population`: a number ID for each site  
- `Individual_ID`: a unique ID for each snail
- `Nematodes`: the number of nematodes found trapped in the shell of the focal snail. NAs are cases where a snail that was collected was accidentally lost during handling

The `colours_raw` dataset is another simple data frame with five columns:

- `Population` and `Individual_ID` as above
- `morph`: the shell colour morph using the traditional notation for _Cepaea_ morphs. Note: band interruptions are not recorded here differently from full bands (they are rare anyway, there might have been 1-5 individuals total, but definitely <10)
- `col`: the background colour as extracted from `morph`
- `nbands`: the number of shell bands as extracted from `morph` (two fused bands count as two bands)
- `fusion`: if the shell has bands, whether or not some of these are fused together

The `points_raw` dataset is a geopackage containing the point locations for each of the 36 visited sites, along with the following information on each site:

- `City`: the urban area to which a site belongs to. For very rural sites, the urban area it is closest to
- `Population`: same as above
- `Date`: the date the site was visited, in dd/mm/yyyy format  
- `N_collectors`: the number of collectors during that visit
- `Sampling_effort_minutes`: total sampling effort **in person-minutes** (i.e., a value of 30 here with `N_collectors`=2 means 2 people searched 15 min each)
- `POP100`, `BUILT100`: the values of population density (`POP`) and built-up surface (`BUILT`) for the 100 x 100 m grid cell the point falls in, from the GHSL rasters (see main text Methods for references). **IMPORTANT**: these are _absolute_ (people, m^2^), not relative (people.km^-2^, % built).  
- `POP1000`, `BUILT1000`: same as above, but using the 1000 m resolution rasters instead
- `SMOD`: same as above, but for the 1000 m resolution SMOD/Degree of Urbanisation raster from the GHSL. This gives values at the finer level 2 of the Degree of Urbanisation classification (https://doi.org/10.2785/706535). We are going to use the coarser level 1 classification into (broadly) highly urban/intermediate/rural and process the data accordingly below.

# Processing data

We extract XY coordinates (for spatial autocorrelation tests) and convert SMOD values to their corresponding upper levels types (see references in main text Methods):

```{r process_spatial-data}
points <- points_raw |> 
  mutate(x=st_coordinates(geom)[,1],
         y=st_coordinates(geom)[,2]) |> 
  mutate(
    SMOD_type =  case_when(SMOD ==30 ~ "3 - Urban Centre",
                      (SMOD >= 20 & SMOD < 30) ~ "2 - Intermediate",
                      (SMOD > 10 & SMOD < 20) ~ "1 - Rural",
                      SMOD == 10 ~ "0 - Water bodies",
                      T ~ NA_character_)
  )
```

We then combine that with the other datasets:

```{r make_final-data}
data <- nematodes_raw |> 
  left_join(colours_raw) |> 
  left_join(points) |> 
  mutate(has.nematodes=Nematodes>0)
```

## Misc numbers

Here we generate some interesting quantities for the Methods section

```{r misc-numbers1}
cat(
  "\n",
  paste0("Number of sites visited: ",dim(points)[1],"\n"),
  paste0("------ of snails: ",dim(data)[1],"\n"), # number of snails
  paste0("------ of sites with snails found: ",length(unique(data$Population)),"\n"),
  paste0("------ of snails scanned for nematodes (i.e. excluding missing snails): ",sum(!is.na(data$Nematodes)),"\n"),
  paste0("------ of sites with snails, after accounting for lost snails: ",length(unique(na.omit(data)$Population)),"\n" ),
  paste0("------ of snails with nematodes: ", sum(data$has.nematodes,na.rm=TRUE), "\n"),
  paste0("------ of nematodes found across all shells: ",sum(data$Nematodes,na.rm=TRUE))
)
```


```{r misc-numbers2}
cat(
  "\n",
  paste0("Collectors per site, mean: ",
         round(mean(points$N_collectors),2),
         ", SD: ",
         round(sd(points$N_collectors),2),
         ", range: ",min(points$N_collectors),"-",max(points$N_collectors), "\n"),
  paste0("Effort in person-minutes per site, mean: ",
         round(mean(points$Sampling_effort_minutes),2),
         ", SD: ",
         round(sd(points$Sampling_effort_minutes),2),
         ", range: ", min(points$Sampling_effort_minutes),"-",max(points$Sampling_effort_minutes))
)
```



```{r misc-numbers3}
table(points$City)  # spread of visited sites between cities

successful_points <- points |> filter(Population %in% data$Population)
table(successful_points$City)    # spread of sites with Cepaea between cities
```



```{r misc-numbers4}
nearest <- st_nearest_feature(successful_points)
nearest_dist <- st_distance(successful_points, successful_points[nearest,], by_element=TRUE)

cat(
  "for sites with snails found, the distance to the nearest site with also snails found \nis on average",
  round(mean(nearest_dist),2),"m \nwith a range of ", paste0(round(min(nearest_dist),2),"-",round(max(nearest_dist),2))
)
```

We can now move to running the models.

# Models

```{r models}
mod_built100 <- glmmTMB(has.nematodes~scale(BUILT100)+City+col+scale(nbands)+fusion+(1|Population),data=data,family=binomial)
mod_built1000 <- glmmTMB(has.nematodes~scale(BUILT1000)+City+col+scale(nbands)+fusion+(1|Population),data=data,family=binomial)
mod_pop100 <- glmmTMB(has.nematodes~scale(POP100)+City+col+scale(nbands)+fusion+(1|Population),data=data,family=binomial)
mod_pop1000 <- glmmTMB(has.nematodes~scale(POP1000)+City+col+scale(nbands)+fusion+(1|Population),data=data,family=binomial)

mod_SMOD <- glmmTMB(has.nematodes~SMOD_type+City+col+nbands+fusion+(1|Population),data=data,family=binomial)

mod_null <- glmmTMB(has.nematodes~City+col+nbands+fusion+(1|Population),data=data,family=binomial)
# we use glmmTMB() rather than lme4::glmer() because some models have convergence issues on the latter
# easy to check that the models with no issues give the same thing (to several decimals) on both packages

```

## testing for model assumption

They're Bernoulli models so in principle shouldn't be any big issue, but let's check anyway:
```{r model-check}
simres <- simulateResiduals(mod_SMOD)
plot(simres)
```

## model comparison

```{r model-selection}
mtable <- model.sel(mod_built100,mod_built1000,mod_pop100,mod_pop1000,mod_SMOD,mod_null,rank="AICc")


tibble(
  model_name = row.names(mtable),
  df=mtable$df,
  logLik=round(mtable$logLik,1),
  AICc = round(mtable$AICc,1),
  delta = round(mtable$delta,2),
  weight = round(mtable$weight,2)
) |> 
  knitr::kable()
```


And there is no spatial autocorrelation to worry about
```{r model-check2}
correlog <- spline.correlog (x = data$x/1000, y = data$y/1000,
                             z = residuals(mod_SMOD, type = "pearson"))

plot(correlog) # no evidence of residual autocorrelation

# calculating per group
groupLocations = data |> 
  as_tibble() |> 
  filter(!is.na(has.nematodes) & !is.na(morph)) |> 
  mutate(residuals=residuals(mod_SMOD, type = "pearson")) |> 
  group_by(x,y,Population) |> 
  summarise(meanres=mean(residuals))

correlog <- spline.correlog (x = groupLocations$x/1000, y = groupLocations$y/1000,
                             z = groupLocations$meanres)

plot(correlog) # no evidence of residual autocorrelation
```

We can move forward with our model then :)

# Model analysis

```{r model-test}

r.squaredGLMM(mod_SMOD)
Anova(mod_SMOD)

contrast(emmeans(mod_SMOD,specs = ~ SMOD_type),method="pairwise")

preds <- emmeans(mod_SMOD,specs = ~ SMOD_type, type="response") |> 
  as_tibble() 
```

# Plot

```{r plot}
aggregated_data <- data |> 
  as_tibble() |> select(-geom) |> 
  group_by(SMOD_type,Population) |>
  summarise(N_nema=sum(has.nematodes,na.rm=TRUE),N=sum(!is.na(has.nematodes)))

ggplot() + 
  geom_point(data=aggregated_data,aes(SMOD_type,N_nema/N,size=N),position = position_jitter(width=0.3,height=0),col="darkgrey") + 
  geom_pointrange(data = preds, aes(x=SMOD_type,y=prob,ymin=asymp.LCL,ymax=asymp.UCL),size=1,pch=21,fill="white")+ 
  scale_size_area("Number of snails") + 
  labs(x="Degree of Urbanisation (GHS-SMOD, 1km grid cell)", y = "Proportion of shells with nematodes")+
  theme_bw()
```
